### 2022-10-05 doc/src/IOMMU.tex
MSI page tables apply to all forms of memory accesses
... Not just naturally aligned 32-bit writes.

Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

\item
an \emph{MSI address mask} and \emph{address pattern}, used together to
recognize certain memory writes from the device as being MSIs; and

@@

The MSI address mask and address pattern are each unsigned integers
with the same width as guest physical page numbers, i.e., 12~bits
narrower than the maximum supported width of a guest physical address.
Their use is explained in Section~\ref{sec:IOMMU-identIncomingMSIs},
``Identification of incoming MSIs for a virtual machine.''

@@

%-----------------------------------------------------------------------
\section{Identification of incoming MSIs for a virtual machine}
\label{sec:IOMMU-identIncomingMSIs}

@@

The MSI address mask and address pattern specified in a device context
(Section~\ref{sec:IOMMU-deviceContexts}) are used to identify the
\mbox{4-KiB} pages of virtual interrupt files in the guest physical
address space of the relevant VM.
An incoming \mbox{32-bit} write made by a device is recognized as an
MSI write to a virtual interrupt file if the destination guest physical
page matches the supplied address pattern in all bit positions that are
zeros in the supplied address mask.
In detail, a write to guest physical address~$A$ is recognized as an
MSI to a virtual interrupt file if
$$
\bigl(\mbox{($A$\,\z{>>}\,12) \z{\&} $\sim$address mask}\bigr)
  = (\mbox{address pattern \z{\&} $\sim$address mask})
$$
where \mbox{\z{>>}\,12} represents shifting right by 12~bits,
an ampersand (\z{\&}) represents bitwise logical AND, and
``$\sim$address mask'' is the bitwise logical complement of the address
mask.

When a write is identified as an MSI from a device, an
\emph{interrupt file number} is extracted from the original guest
physical address as
\begin{displayLinesTable}
interrupt file number = extract($A$\,\z{>>}\,12, address mask) \\
\end{displayLinesTable}
Here, extract($x$,~$y$) is a ``bit extract'' that discards all bits from
$x$ whose matching bits in the same positions in the mask~$y$ are
zeros, and packs the remaining bits from $x$ contiguously at the
least-significant end of the result, keeping the same bit order as $x$
and filling any other bits at the most-significant end of the result
with zeros.
For example, if the bits of $x$ and $y$ are
\begin{displayLinesTable}[r@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c]
$x$ = & a & b & c & d & e & f & g & h \\
$y$ = & 1 & 0 & 1 & 0 & 0 & 1 & 1 & 0 \\
\end{displayLinesTable}
then the value of extract($x$,~$y$) has bits 0~0~0~0~a~c~f~g.

@@

Whenever an \mbox{IOMMU} recognizes an incoming write from a device as
an MSI by the method specified in the previous section, the MSI is
translated or converted by consulting the MSI page table configured for
the device, instead of using the regular translation data structures
that apply to all other memory accesses from the same device.

Only naturally aligned \mbox{32-bit} writes from a device are possible
MSIs.
For other forms of memory accesses by a device (such as reads, writes
of other sizes, or misaligned writes), the regular translation data
structures are always applied, even if the address matches that of a
proper MSI.

An MSI page table is a flat array of MSI page table entries (MSI PTEs),
each 16~bytes.
MSI page tables have no multi-level hierarchy like regular {\RISCV}
page tables do.
Rather, every MSI PTE is a leaf entry specifying the translation or
conversion of writes made to a particular \mbox{4-KiB} guest physical
page that a virtual interrupt file occupies (or may occupy) in the
relevant virtual machine.
To select an individual MSI PTE from an MSI page table, the PTE array
is indexed by the interrupt file number extracted from the destination
guest physical address of the incoming MSI write by the formula of the
previous section.
Each MSI PTE may specify either the address of a real guest interrupt
file that substitutes for the targeted virtual interrupt file (as in
Figure~\ref{fig:IOMMU-guestIntrFiles}), or a memory-resident interrupt
file in which to store incoming MSIs for the virtual interrupt file
(as in Figure~\ref{fig:IOMMU-MRIF}).

@@

If V~=~1 and the custom-use bit C~=~0, then bit~2 of the first
doubleword is field~W (Write-through).
If W~=~1, the MSI PTE specifies \emph{write-through mode} for incoming
MSIs, and if W~=~0, it specifies \emph{MRIF mode}.
The interpretation of an MSI PTE for each of these two modes is
detailed further in the next two subsections.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{MSI PTE, write-through mode}

When an MSI PTE has fields V~=~1, C~=~0, and W~=~1
(write-through mode), the PTE's complete format is:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l@{\quad}l]
First doubleword:  & bit 63     & C, = 0 \\
                   & bits 53:10 & PPN \\
                   & bit 2      & W, = 1 \\
                   & bit 0      & V, = 1 \\
\noalign{\medskip}
Second doubleword: & ignored \\
\end{displayLinesTable}
All other bits of the first doubleword are reserved and must be set to
zeros by software.
The second doubleword is ignored by an \mbox{IOMMU} so is free for software
to use.

An incoming MSI write is translated by replacing the write's original
address bits 12 and above (the guest physical page number) with field
PPN (Physical Page Number) from the PTE, while retaining the original
address bits 11:0 (the page offset).
This translated address is either zero-extended or clipped at the upper
end as needed to make it the width of a real physical address for the
machine.
The original memory write from the device is then passed onward to the
memory system with the new address.

An MSI PTE in write-through mode allows a hypervisor to route an
MSI intended for a virtual interrupt file to go instead to a guest
interrupt file of a real IMSIC in the machine.

@@

For RV64, the first doubleword of an MSI PTE in write-through mode
has the same encoding as a regular {\RISCV} leaf PTE for Sv39, Sv48,
Sv39x4, or Sv48x4 page-based address translation, with PTE fields D, A,
G, U, X, and R all zeros and W~=~1.
Hence, the MSI PTE's first doubleword appears the same as a regular
PTE that grants write permission (W~=~1) but not read or execute
permissions (X~= R =~0).
This same-encoded regular PTE would translate an MSI write the same as
the actual MSI PTE, except that what would be the PTE's accessed (A),
dirty (D), and user (U) bits are all zeros.
An \mbox{IOMMU} needs to treat only these three bits differently for an MSI
PTE versus a regular RV64 leaf PTE.

@@

If memory-resident interrupt files are supported and an MSI PTE has
fields V~=~1, C~=~0, and W~=~0 (MRIF mode), the PTE's complete format
is:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l@{\quad}l]
First doubleword:  & bit 63     & C, = 0 \\
                   & bits 53:7  & MRIF Address[55:9] \\
                   & bit 2      & W, = 0 \\
                   & bit 0      & V, = 1 \\
\noalign{\medskip}
Second doubleword: & bit 60     & NID[10] \\
                   & bits 53:10 & NPPN \\
                   & bits 9:0   & NID[9:0] \\
\end{displayLinesTable}
All other PTE bits are reserved and must be set to zeros by software.

@@

@@

If the incoming MSI write is to guest physical address~$A$ and its
\mbox{32-bit} data is~$D$ when interpreted in the byte order indicated
by bit~2 of $A$, then the MSI is processed by the \mbox{IOMMU} as follows:
If either $A$\z{[}11:3\z{]} or $D$\z{[}31:11\z{]} is not zero, or if
bit~2 of $A$ is one and big-endian MSIs are not supported, then the
incoming write is accepted but discarded.
Else, the original incoming write is replaced by this sequence:
\begin{enumerate}

@@

\begin{commentary}
While \mbox{IOMMU}s are expected typically to cache MSI PTEs that are
configured in write-through mode (W~=~1), they might not cache PTEs
configured in MRIF mode (W~=~0).
Two reasons together justify not caching MSI PTEs in MRIF mode:
First, the information and actions required to store an MSI to an MRIF
are far different than normal address translation; and
second, by their nature, MSIs to MRIFs should occur less frequently.
Hence, an \mbox{IOMMU} might perform MRIF-mode processing solely as
an extension of cache-miss page table walks, leaving its address
translation cache oblivious to MRIF-mode MSI PTEs.
\end{commentary}

>>>>

\item
an \emph{MSI address mask} and \emph{address pattern}, used together to
identify pages in the guest physical address space
that are the destinations of MSIs; and

@@

The MSI address mask and address pattern are each unsigned integers
with the same width as guest physical page numbers, i.e., 12~bits
narrower than the maximum supported width of a guest physical address.
Their use is explained in Section~\ref{sec:IOMMU-identIncomingMSIs},
``Identification of page addresses of a VM's interrupt files.''

@@

%-----------------------------------------------------------------------
\section{Identification of page addresses of a VM's interrupt files}
\label{sec:IOMMU-identIncomingMSIs}

@@

The MSI address mask and address pattern specified in a device context
(Section~\ref{sec:IOMMU-deviceContexts}) are used to identify the
\mbox{4-KiB} pages of virtual interrupt files in the guest physical
address space of the relevant VM.
An incoming \mbox{32-bit} write made by a device is recognized as an
MSI write to a virtual interrupt file if the destination guest physical
page matches the supplied address pattern in all bit positions that are
zeros in the supplied address mask.
In detail, a memory access to guest physical address~$A$ is
an access to a virtual interrupt file's memory-mapped page if
$$
\bigl(\mbox{($A$\,\z{>>}\,12) \z{\&} $\sim$address mask}\bigr)
  = (\mbox{address pattern \z{\&} $\sim$address mask})
$$
where \mbox{\z{>>}\,12} represents shifting right by 12~bits,
an ampersand (\z{\&}) represents bitwise logical AND, and
``$\sim$address mask'' is the bitwise logical complement of the address
mask.

When a memory access is found to be to a virtual interrupt file, an
\emph{interrupt file number} is extracted from the original guest
physical address as
\begin{displayLinesTable}
interrupt file number = extract($A$\,\z{>>}\,12, address mask) \\
\end{displayLinesTable}
Here, extract($x$,~$y$) is a ``bit extract'' that discards all bits from
$x$ whose matching bits in the same positions in the mask~$y$ are
zeros, and packs the remaining bits from $x$ contiguously at the
least-significant end of the result, keeping the same bit order as $x$
and filling any other bits at the most-significant end of the result
with zeros.
For example, if the bits of $x$ and $y$ are
\begin{displayLinesTable}[r@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c@{ }c]
$x$ = & a & b & c & d & e & f & g & h \\
$y$ = & 1 & 0 & 1 & 0 & 0 & 1 & 1 & 0 \\
\end{displayLinesTable}
then the value of extract($x$,~$y$) has bits 0~0~0~0~a~c~f~g.

\clearpage

@@

When an \mbox{IOMMU} determines that a memory access is to a virtual
interrupt file as specified in the previous section, the access is
translated or converted by consulting the MSI page table configured for
the device, instead of using the regular translation data structures
that apply to all other memory accesses from the same device.

An MSI page table is a flat array of MSI page table entries (MSI PTEs),
each 16~bytes.
MSI page tables have no multi-level hierarchy like regular {\RISCV}
page tables do.
Rather, every MSI PTE is a leaf entry specifying the translation or
conversion of accesses made to a particular \mbox{4-KiB} guest physical
page that a virtual interrupt file occupies (or may occupy) in the
relevant virtual machine.
To select an individual MSI PTE from an MSI page table, the PTE array
is indexed by the interrupt file number extracted from the destination
guest physical address of the incoming memory access by the formula of the
previous section.
Each MSI PTE may specify either the address of a real guest interrupt
file that substitutes for the targeted virtual interrupt file (as in
Figure~\ref{fig:IOMMU-guestIntrFiles}), or a memory-resident interrupt
file in which to store incoming MSIs for the virtual interrupt file
(as in Figure~\ref{fig:IOMMU-MRIF}).

@@

If V~=~1 and the custom-use bit C~=~0, then bits~2:1 of the first
doubleword contain field~M (Mode).
If M~=~3, the MSI PTE specifies \emph{pass-through mode} for
accesses to the page, and if M~=~1, it specifies \emph{MRIF mode}.
Values of 0 and 2 for M are reserved.
The interpretation of an MSI PTE for each of the two defined modes is
detailed further in the next two subsections.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{MSI PTE, pass-through mode}

When an MSI PTE has fields V~=~1, C~=~0, and M~=~3
(pass-through mode), the PTE's complete format is:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l@{\quad}l]
First doubleword:  & bit 63     & C, = 0 \\
                   & bits 53:10 & PPN \\
                   & bits 2:1   & M, = 3 \\
                   & bit 0      & V, = 1 \\
\noalign{\medskip}
Second doubleword: & ignored \\
\end{displayLinesTable}
All other bits of the first doubleword are reserved and must be set to
zeros by software.
The second doubleword is ignored by an \mbox{IOMMU} so is free for software
to use.

A memory access within the page covered by the MSI PTE
is translated by replacing the access's original
address bits 12 and above (the guest physical page number) with field
PPN (Physical Page Number) from the PTE, while retaining the original
address bits 11:0 (the page offset).
This translated address is either zero-extended or clipped at the upper
end as needed to make it the width of a real physical address for the
machine.
The original memory access from the device is then passed onward to the
memory system with the new address.

An MSI PTE in pass-through mode allows a hypervisor to route an
MSI write intended for a virtual interrupt file to go instead to a guest
interrupt file of a real IMSIC in the machine.

@@

For RV64, the first doubleword of an MSI PTE in pass-through mode
has the same encoding as a regular {\RISCV} leaf PTE for Sv39, Sv48,
Sv39x4, or Sv48x4 page-based address translation, with PTE fields D, A,
G, U, and X all zeros and W~= R~=~1.
Hence, the MSI PTE's first doubleword appears the same as a regular
PTE that grants read and write permission (R~= W~=~1) but not execute
permissions (X~=~0).
This same-encoded regular PTE would translate an MSI write the same as
the actual MSI PTE, except that what would be the PTE's accessed (A),
dirty (D), and user (U) bits are all zeros.
An \mbox{IOMMU} needs to treat only these three bits differently for an MSI
PTE versus a regular RV64 leaf PTE.

@@

If memory-resident interrupt files are supported and an MSI PTE has
fields V~=~1, C~=~0, and M~=~1 (MRIF mode), the PTE's complete format
is:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l@{\quad}l]
First doubleword:  & bit 63     & C, = 0 \\
                   & bits 53:7  & MRIF Address[55:9] \\
                   & bits 2:1   & M, = 1 \\
                   & bit 0      & V, = 1 \\
\noalign{\medskip}
Second doubleword: & bit 60     & NID[10] \\
                   & bits 53:10 & NPPN \\
                   & bits 9:0   & NID[9:0] \\
\end{displayLinesTable}
All other PTE bits are reserved and must be set to zeros by software.

@@

Memory accesses by I/O devices to addresses within a page
covered by an MRIF-mode PTE are handled by the IOMMU
instead of being passed through to the memory system.
For a read access, the IOMMU returns zero as the read value.
A write access is either interpreted as an MSI, resulting
in an update of the destination MRIF, or is discarded.

@@

If an incoming write access is for a data size other
than 32~bits or the address is not \mbox{32-bit}-aligned,
then the write is accepted by the IOMMU but discarded.
Otherwise, if the write is to guest physical address~$A$ and its
\mbox{32-bit} data is~$D$ when interpreted in the byte order indicated
by bit~2 of $A$, then the write is processed as follows:
If either $A$\z{[}11:3\z{]} or $D$\z{[}31:11\z{]} is not zero, or if
bit~2 of $A$ is one and big-endian MSIs are not supported, then the
incoming write is accepted but discarded.
Else, the original incoming write is recognized
as an MSI and is replaced by this sequence:
\begin{enumerate}

@@

\begin{commentary}
While \mbox{IOMMU}s are expected typically to cache MSI PTEs that are
configured in pass-through mode (M~=~3), they might not cache PTEs
configured in MRIF mode (M~=~1).
Two reasons together justify not caching MSI PTEs in MRIF mode:
First, the information and actions required to store an MSI to an MRIF
are far different than normal address translation; and
second, by their nature, MSIs to MRIFs should occur less frequently.
Hence, an \mbox{IOMMU} might perform MRIF-mode processing solely as
an extension of cache-miss page table walks, leaving its address
translation cache oblivious to MRIF-mode MSI PTEs.
\end{commentary}
